---
title: "Reproducible research: Project1"
author: "Nadia Stavisky"
date: "7/21/2019"
output: html_document
---

```{r global.options, include = FALSE, echo=FALSE}
#load library 'knitr', set a golobal default options:
library(knitr)
library(ggplot2)
library(formatR)
library(dplyr)
opts_chunk$set(
    echo = TRUE, # if FALSE knitr will not display code in the code chunk above it's results
    cache = FALSE, # if TRUE knitr will cache the results to reuse in future knits
    fig.width   = 10, # the width for plots created by code chunk
    fig.height  = 8, # the height for plots created by code chunk
    fig.align   = 'center', # how to align graphics in the final doc. 'left', 'right', 'center'
    fig.path    = 'figs/',  # file path to the directory where knitr shall store the graphics files
    results     = 'asis',   # knitr will pass through results without reformatting them
    tidy = TRUE,
    tidy.opts=list(width.cutoff=60),
    strip.white = TRUE)     # if FALSE knitr will not remove white spaces at the beg or end of code chunk
```

##Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](https://www.fitbit.com/home), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone Up](https://jawbone.com/up). These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Loading and preprocessing the data

The data for this assignment downloaded from the course web site:

Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity)
```{r}
#download data, unzip, import file into R object
datapath <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(datapath, destfile = "./Fdata.zip")
unzip("Fdata.zip")
activity <- read.csv2("activity.csv", header = TRUE, sep = ",")
```
The variables included in this dataset are:
```{r}
names(activity)
```
data set dimentions:
```{r}
dim(activity)
```

transform date field from factors into dates:
```{r}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

structure of the dataset:
```{r}
str(activity)
```

dataset summary:
```{r}
summary(activity)
```

quick view on data set:
```{r}
head(activity)
```


##What is mean total number of steps taken per day?
###For this part of the assignment, the missing values in the dataset were ignored.

The total number of steps taken per day:
```{r}
ttl_steps <- activity %>% 
  filter(complete.cases(activity)) %>%
  group_by(date) %>%
  summarise(steps_sum = sum(steps, na.rm = TRUE))
head(ttl_steps)
summary(ttl_steps)
```
A histogram of the total number of steps taken each day

```{r}
a <- ggplot(data = ttl_steps, aes(ttl_steps$steps_sum))
a + geom_histogram(breaks = seq(0,25000, by = 2500), col="blue", fill="blue", alpha=.4) +
   labs(title = "Histogram of Total Number of Steps Each Day", x = "Number of Steps") +
   theme_bw()
```
```{r}
mean <- mean(ttl_steps$steps_sum)
median <- median(ttl_steps$steps_sum)
```
the mean and median of the total number of steps taken per day are `r mean` and `r median` correspondingly.

##What is the average daily activity pattern?

The verage daily activity pattern by interval:
```{r}
avg_steps <- activity %>% 
  filter(complete.cases(activity)) %>%
  group_by(interval) %>%
  summarise(steps_avg = mean(steps, na.rm = TRUE))
head(avg_steps)
summary(avg_steps)
max <- as.numeric(c(avg_steps[which.max(avg_steps$steps_avg),1]))

```

A time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis):

```{r}
b <- ggplot(data = avg_steps, aes (x = as.integer(avg_steps$interval), y = avg_steps$steps_avg))
b + geom_line() +
  geom_vline(xintercept = max, color = "blue", size=1)

```

the `r max` 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

##Imputing missing values
```{r}
summary(activity)
notcomplete <- sum(!complete.cases(activity))
```
Note that there are a number of days/intervals where there are missing values (coded as \color{red}{\verb|NA|}NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

There are `r notcomplete` total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs).

We used median for that day as a strategy for filling in all of the missing values in the dataset.

The median number of steps by date:
```{r}
mdn_steps <- activity %>% 
  
  group_by(date) %>%
  summarise(steps_mdn = median(steps, na.rm = TRUE))
head(mdn_steps)
summary(mdn_steps)
```
 
```{r}
activity2 <- activity %>%
  group_by(date) %>%
  mutate(steps = replace(steps, is.na(steps), median(steps, na.rm = TRUE)))
```

```{r}
summary(activity2)
notcomplete2 <- sum(!complete.cases(activity2))
```
The total number of steps taken per day:
```{r}
ttl_steps2 <- activity2 %>% 
  group_by(date) %>%
  summarise(steps_sum = sum(steps))
head(ttl_steps2)
summary(ttl_steps2)
```

A histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}
c <- ggplot(data = ttl_steps, aes(ttl_steps$steps_sum))
c + geom_histogram(breaks = seq(0,25000, by = 2500), col="blue", fill="blue", alpha=.4) +
   labs(title = "Histogram of Total Number of Steps Each Day", x = "Number of Steps") +
   theme_bw()
```


Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
Are there differences in activity patterns between weekdays and weekends?
For this part the \color{red}{\verb|weekdays()|}weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
